---
title: "506 HW1 1"
author: "Steven Leone"
format: html
editor: visual
---

This is the link to the repo: <https://github.com/leonestv-byte/506-hw1>.

This is the file for DATASCI 506 HW1.

```{r}
#| echo: false
# install.packages("readODS")
library(readODS)
library(dplyr)

```

## Exercise 1

a)  Import the data into a data.frame in R. Use the dataset documentation to assign appropriate variable names. (Note: Downloading and unzipping the file can take place outside of your submitted document, but importing the file should be in the submission.)

```{r}
df <- read.table("data.csv", sep =",", header = TRUE)

descriptions <- read_ods("cultivars-description.ods")

```

b)  Using the information provided in the dataset documentation: Verify that there are 320 observations in the dataset; 40 distinct cultivars; and all the observations were collected from two seasons. Produce a table showing the number of observations from each season.

```{r}
print(paste("The number of observations in the dataset is: ", nrow(df)))
```

```{r}
print(paste("The number of distinct cultivars is: ", nrow(descriptions)))
```

c)  Use the data to answer the following questions:

What is the correlation between plant height (PH) and grain yield (GY)?

```{r}
cor(df$PH, df$GY)
```

Calculate the correlation between PH and GY separately for each season.

```{r}
season_1_cor <- cor(df[df$Season == 1, ]$PH, df[df$Season == 1, ]$GY,)
season_1_cor

```

```{r}
season_2_cor <- cor(df[df$Season == 2, ]$PH, df[df$Season == 2, ]$GY,)
season_2_cor

```

Which season has the higher correlation?

```{r}
season_with_higher_correlation <-ifelse(abs(season_1_cor) > abs(season_2_cor), "Season 1", "Season 2")
print(season_with_higher_correlation)
```

What is the grain yield of the soybean observation with the greatest plant height?

```{r}
df[df$PH == max(df$PH), ]$GY
```

What percentage of observations have a thousand seed weight (MHG) greater than the overall average thousand seed weight?

```{r}
nrow(df[df$MHG > mean(df$MHG), ]) / nrow(df)

```

Which cultivar has the highest average grain yield?

```{r}
result <- df |>
  group_by(Cultivar) |>       
  summarize(
    avg_value = mean(GY)     
  )

result[result$avg_value == max(result$avg_value), ]$Cultivar
```

d)  Create a table showing the average value of each quantitative variable:

One row containing overall averages.

```{r}
averages_table <- data.frame(
  Repetition = mean(df$Repetition, na.rm = TRUE),
  IFP        = mean(df$IFP, na.rm = TRUE),
  PH         = mean(df$PH, na.rm = TRUE),
  
NLP = mean(df$NLP, na.rm = TRUE),

NGP = mean(df$NGP, na.rm = TRUE),

NGL = mean(df$NGL, na.rm = TRUE),

NS = mean(df$NS, na.rm = TRUE),

MHG = mean(df$MHG, na.rm = TRUE),

GY = mean(df$GY, na.rm = TRUE)

)



averages_table
```

One row for each season containing season-specific averages.

```{r}
df_season_1 = df[df$Season == 1,]
df_season_2 = df[df$Season == 2,]

averages_table_one <- data.frame(
  Repetition = mean(df_season_1$Repetition, na.rm = TRUE),
  IFP        = mean(df_season_1$IFP, na.rm = TRUE),
  PH         = mean(df_season_1$PH, na.rm = TRUE),
  
NLP = mean(df_season_1$NLP, na.rm = TRUE),

NGP = mean(df_season_1$NGP, na.rm = TRUE),

NGL = mean(df_season_1$NGL, na.rm = TRUE),

NS = mean(df_season_1$NS, na.rm = TRUE),

MHG = mean(df_season_1$MHG, na.rm = TRUE),

GY = mean(df_season_1$GY, na.rm = TRUE)

)



averages_table_one




averages_table_two <- data.frame(
  Repetition = mean(df_season_2$Repetition, na.rm = TRUE),
  IFP        = mean(df_season_2$IFP, na.rm = TRUE),
  PH         = mean(df_season_2$PH, na.rm = TRUE),
  
NLP = mean(df_season_2$NLP, na.rm = TRUE),

NGP = mean(df_season_2$NGP, na.rm = TRUE),

NGL = mean(df_season_2$NGL, na.rm = TRUE),

NS = mean(df_season_2$NS, na.rm = TRUE),

MHG = mean(df_season_2$MHG, na.rm = TRUE),

GY = mean(df_season_2$GY, na.rm = TRUE)

)



averages_table_two
```

e)  Carry out a series of t-tests to examine whether the ngrain yield (GY) differs between the two seasons. State the null and the alternative hypotheses. Present the R output and interpret the results.

alternative hypothesis: alternative hypothesis: true difference in means between group 1 and group 2 is not equal to 0

null hypothesis: alternative hypothesis: true difference in means between group 1 and group 2 is equal to 0

```{r}
t_test_result <- t.test(GY ~ Season, data = df)

print(t_test_result)
```

We fail to reject the null. There is no statistically significant difference.

## Exercise 2

a)  Import the data into a data.frame in R. As with the previous data, you may download the data outside of your submission, but importation should take place inside the problem set submission.

```{r}


adult_df <- read.csv(
  file = "adult.data", 
  header = FALSE, 
  strip.white = TRUE,  
  na.strings = "?"   
)

head(adult_df)
```

b)  Clean and simplify the variable names.

```{r}
col_names <- c("age", "workclass", "fnlwgt", "education", "education_num", "marital_status", "occupation", "relationship", "race", "sex", "capital_gain", "capital_loss", "hours_per_week", "native_country", "income")

adult_df <- read.csv(
  file = "adult.data", 
  header = FALSE, 
  col.names = col_names, 
  strip.white = TRUE,  
  na.strings = "?"    
)

head(adult_df)

```

c)  Examine missing values. Report the number of missing observations for each variable.

```{r}
print(paste("The number of rows with missing workclass is", nrow(adult_df[is.na(adult_df$workclass), ])))
print(paste("The number of rows with missing age is",nrow(adult_df[is.na(adult_df$age), ])))
print(paste("The number of rows with missing fnlwgt is",nrow(adult_df[is.na(adult_df$fnlwgt), ])))
print(paste("The number of rows with missing education is",nrow(adult_df[is.na(adult_df$education), ])))
print(paste("The number of rows with missing education number is",nrow(adult_df[is.na(adult_df$education_num), ])))
print(paste("The number of rows with missing marital status is",nrow(adult_df[is.na(adult_df$marital_status), ])))
print(paste("The number of rows with missing occupation is",nrow(adult_df[is.na(adult_df$occupation), ])))
print(paste("The number of rows with missing relationship is",nrow(adult_df[is.na(adult_df$relationship), ])))
print(paste("The number of rows with missing race is",nrow(adult_df[is.na(adult_df$race), ])))
print(paste("The number of rows with missing sex is",nrow(adult_df[is.na(adult_df$sex), ])))
print(paste("The number of rows with missing capital gain is",nrow(adult_df[is.na(adult_df$capital_gain), ])))
print(paste("The number of rows with missing capital loss is",nrow(adult_df[is.na(adult_df$capital_loss), ])))
print(paste("The number of rows with missing hours per week is",nrow(adult_df[is.na(adult_df$hours_per_week), ])))
print(paste("The number of rows with missing native country is",nrow(adult_df[is.na(adult_df$native_country), ])))
print(paste("The number of rows with missing income is",nrow(adult_df[is.na(adult_df$income), ])))
```

d)  Restrict the sample to individuals working at least 10 hours per week. Confirm the sample size before and after filtering.

```{r}
nrow(adult_df)

adult_df = adult_df[adult_df$hours_per_week >= 10,]

nrow(adult_df)
```

e)  Remove observations with missing values for occupation, workclass, or native country. Explain why.

These are the only columns with missing data.

```{r}

adult_df <- adult_df[!is.na(adult_df$occupation),]

adult_df <- adult_df[!is.na(adult_df$workclass),]

adult_df <- adult_df[!is.na(adult_df$native_country),]
```

f)  Eliminate any observations with: age less than 18, age greater than 90, education level identified as Preschool.

```{r}

adult_df <- adult_df[adult_df$age >= 18,]

adult_df <- adult_df[adult_df$age <= 90,]

adult_df <- adult_df[!adult_df$education != 'Preschool',]
```

g)  Report your final sample size.

```{r}
nrow(adult_df)
```

## Exercise 3

a)  Write a function nextHappy() that given a positive integer, computes the next value in the happy-number sequence. Be sure to provide a reasonable error on an invalid input. Be sure to document your function (see instructions above).

```{r}
nextHappy <- function(input) {
  
  digits <- as.numeric(strsplit(as.character(input), "")[[1]])
  number <- 0
  for (digit in digits) {
    number <- number + (digit * digit)
  }
  return(number)
}

```

Example 1:

```{r}
print(nextHappy(19))
```

Example 2:

```{r}
print(nextHappy(82))
```

b)  Create a function happySequence() that returns the sequence beginning at a given number. Use your nextHappy() function to perform the calculation. Be sure to provide a reasonable error on an invalid input. Be sure to document your function (see instructions above).

```{r}
happySequence <- function(input) {
  seen <- c()
  curr <- input
  sequence <- c(input)
  
  while(curr != 1 && !(curr %in% seen)) {
    seen <- c(seen, curr)
    curr <- nextHappy(curr)
    sequence <- c(sequence, curr)
    
  }
  
  is_happy <- (curr == 1)
  
  if (is_happy) {
    return(list(
      sequence = sequence,
      isHappy = is_happy
      ))
  } else {
    return(list(
      sequence = c(),
      isHappy = is_happy
      ))
  }

}


```

```{r}
print(happySequence(7))
```

```{r}
print(happySequence(4))
```

c)  

```{r}
count <- 0

for (i in 1:100) {
  is_happy <- happySequence(i)$isHappy
  if (is_happy) {count <- count + 1}
}

print(count)
print(count/100)

ans <- 15999
print(ans)
```
